# Applied Bayesian Analysis 

## Practical Class 1: Conjugate Bayesian Inference using R and Learn-Bayes

## 1. Surgery example from lectures

Load the relevant libraries used throughout this analysis.
```{r, results="hide", message=FALSE}
library(LearnBayes)
library(pscl)
library(ggplot2)
```

Plot the prior distribution.
```{r warning=FALSE}
prior <- c(3,27)
xs <- seq(0,1,length=100)
ys <- dbeta(xs,3,27)
qplot(xs, ys, geom="line")
```

Find the 95% symmetric confidence interval.
```{r}
int1 <- qbeta(c(0.025,0.975),3,27) # symmetric 95% interval
int1
width1 <- int1[2]-int1[1]
```

Find a 95% highest density interval.
```{r warning=FALSE}
int2 <- betaHPD(3,27, 0.95, plot=TRUE)
int2
width2 <- int2[2]-int2[1]
```

Reduction in width of interval with HPD.
```{r}
100*(width1-width2)/width1 #  % 
```

Probability the true mortality risk is greater than 0.2.
```{r}
1-pbeta(0.2,3,27)        #upper tail area
```

What is the predictive probability of observing at least 6 deaths in the next 20 patients?
```{r}
pbetap(prior,20,0:20)   # beta-binomial predictive distribution
sum(pbetap(prior,20,6:20))   # sum to give upper tail area
predplot(prior,20,6 )
```

We do 10 operations and all the patients survive. Plot the prior, likelihood, and posterior distribution for the risk.
```{r}
data <- c(0,10) # no. "successes, failures" (note that here 'failure' is a survival!)
triplot(prior,data) # posterior is beta[3,37]
```

What was our prior predictive probability for observing so few deaths?
```{r}
pbetap(prior,20,0)    # prob no deaths
```

What is the posterior mean and 95% highest posterior density interval for the mortality
risk?
```{r, warning=FALSE}
betaHPD(3,37, 0.95, plot=TRUE) # seem to need to calculate posterior parameters yourself
```

What is the posterior probability that the true mortality rate is greater than 0.2?
```{r}
1-pbeta(0.2,3,27) # tail area of posterior
```

Suppose 20 more patients will be operated upon. What is the probability that at least
2 of them die?
```{r}
post <- prior+data  # posterior is beta[3,37]
sum(pbetap(post,20,2:20)) # upper tail area
```

## 2. Is the coin biased? Bayes with a discrete prior distribution

```{r}
p <- c(0.0,0.1,.2,.3,.4,.5,.6,.7,.8,.9,1.0) # possible values of the 'chance'
prob <- c(0.02,0.02,0.02,0.02,0.02,0.80,0.02,0.02,0.02,0.02,0.02) # prior distribution
n <- 10
s <- 0:10
y <- pdiscp(p,prob,n,s) # predictive distribution for a discrete prior (mixture of binomials)
qplot(s,y)  # crude way of plotting predictive distribution
```
Loosely, 0 heads is more likely than 1 head, as if $\theta=0$ then certain to get 0 heads, and may get 0 even if $\theta > 0$.

Probability of 8 or more heads.
```{r}
sum(pdiscp(p,prob,n,8:10))
```

You flip the coin 10 times and it comes down heads in 8 flips. What is your posterior
probability that the coin is unbiased ($\theta = 0.5$)?
```{r}
data <- c(8,2)  # 8 successes and 2 failures
y <- pdisc(p,prob,data)  # posterior distribution with discrete prior
qplot(p,y)  # crude way of plotting posterior distribution
y[6]  # posterior probability of a fair (unbiased) coin
```

## 3. Drug

What is the posterior mean and 95% highest density interval for the response rate?
```{r, warning=FALSE}
m <- 0.4 # mean
v <- 0.1*0.1 # variance
a  <-  m*( m*(1-m)/v - 1) # prior 'number of successes'
a
b  <-  (1-m)*( m*(1-m)/v -1) # prior 'number of failures'
b
prior <- c(a,b)
data <- c(15,5) # 15 successes and 5 failures
post <- prior+data # posterior parameters
post
post[1]/(post[1]+post[2])  # posterior mean (properties of beta distribution)
betaHPD(24.2,18.8, 0.95, plot=TRUE)
```

Examine the prior and likelihood.
```{r}
triplot(prior,data)
```

What was the prior predictive probability of observing at least 15 successes?
```{r}
sum(pbetap(prior,20,15:20)) #  prior predictive probability of getting at least 15 successes
```

```{r}
m <- 0.8  # alternative prior for 'winners'
v <- 0.1*0.1
a2  <-  m*( m*(1-m)/v - 1) #
a2
b2  <-  (1-m)*( m*(1-m)/v -1) #
b2
```

Plot this 'mixture' prior distribution
```{r}
x <- seq(0, 1, len=100)
y <- 0.95*dbeta(x,a,b)+0.05*dbeta(x,a2,b2)
qplot(x, y, geom="line") # mixture prior distribution
```

What is the posterior probability that the drug is a 'winner'?
```{r}
probs <- c(.95, .05)  # mixing probs
beta.par1 <- c(a,b)  # beta params for 'main prior' (model 1)
beta.par2 <- c(a2,b2) #beta params for 'main prior' (model 2)
betapar <- rbind(beta.par1,beta.par2) # create matrix of beta-parameters
binomial.beta.mix(probs,betapar,data) # function that outputs posterior weights on two 'models', and within-model posterior parameters
0.95*sum(pbetap(beta.par1,20,15:20)) + 0.05*sum(pbetap(beta.par2,20,15:20))# predictive tail-area
```

## 4. Systolic blood pressure: normal-normal analysis

```{r}
# (there are probably functions that can do all this)
priormean <- 120
priorsd <- 10
x <- seq(80, 160, len=100)
y <- dnorm(x,mean=priormean,sd=priorsd)
qplot(x, y, geom="line")
```

What is the posterior mean and 95% HDI for the woman's SBP?
```{r}
n <- 2
ybar <- 130
sigma <- 5
n0 <- (sigma/priorsd)^2
postmean <-  (n*ybar + n0*priormean)/(n+n0)
postmean
postprec <- (n+n0)/sigma^2
postsd <-  1/sqrt(postprec) # precision = 1/SD^2
HPDint  <-  postmean + qnorm(c(0.025,0.975))*postsd # +/- 1.96
HPDint
```

How does this answer compare with a maximum likelihood solution
```{r}
# MLE corresponds to n0=0
HPDmle  <-  ybar + qnorm(c(0.025,0.975))*sigma/sqrt(n) # +/- 1.96
HPDmle
```

## 5. Leukaemia: Poisson-Gamma analysis

Show that the appropriate conjugate prior is a gamma distribution with mean 1 and
standard deviation 0.5.
```{r}
#conjugate prior for Poisson is Gamma
mean <- 1
sd <- 0.5
# mean = a/b; sd = sqrt(a)/b, so
a  <-  (mean/sd)^2
a
b  <-  a/mean
b
```

```{r}
x <- seq(0, 4, len=100)
y <- dgamma(x,4,rate=4)
qplot(x, y, geom="line") # care with parameterisation of gamma!
```

Using the ideas of prior/data compatibility, what is the predictive probability of getting
such an extreme result, given the prior distribution?
```{r}
# predictive distribution for number of cases is negative binomial
# but remember that now have a Poisson mean of theta*E
# and so cannot just use formula on slide 2-12
E <- 0.25
y <- 0:10
preddens <-  E^y* gamma(a+y)/(gamma(a)*gamma(y+1))*b^a/(b+E)^(a+y)
qplot(y,preddens)
pvalue <- sum(preddens[4:10]) # pred prob of getting at least 4
pvalue
```

Suppose this area had been identified as being the most extreme out of 200 regions,
how might that change your conclusions?
```{r}
# probability of getting such an extreme result in 200 separate regions
1 - (1-pvalue)^200
```

If we accept the compatibility of the data with the prior, what is a 95% region for the
SIR?
```{r}
apost <- a+4
bpost <- b+E
x <- seq(0, 4, len=100)
y <- dgamma(x,apost,rate=bpost)
qplot(x, y, geom="line")
qgamma(c(0.025,0.975),apost,rate=bpost)
```

Suppose we accept a 1.5-fold excess risk of leukemia around the nuclear processing plant to be a cause for concern.  To what extend does the posterior support values for $\theta > 1.5$.
```{r}
1-pgamma(1.5,apost,rate=bpost)
```